Exploring Molecular Conformations and Related Time Scales via Nonequilibrium Molecular Dynamics Simulation

### 2013-08-06 writedown/npt-neq/alanine.tex
more revises

\title{Exploring Molecular Conformations and Related Time Scales via Nonequilibrium Molecular Dynamics Simulation}
% \title{The non-equilibrium molecular dynamics simulation of the alanine dipeptide under electric field}
\author{Han Wang}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
% \affiliation{Zuse Institute Berlin, Germany}
\author{Luigi Delle Site}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

The possible effect of the electric field (EF) on the conformation of
proteins gains increasing attention recently, because it closely
related to the health problem of exposing human tissue in the
electromagnetic radiation. Despite the importance of this problem, a
complete understanding of how the electric field influence the  protein
is still lacking, therefore, a considerable number of
experimental~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects} studies have been devoted to this topic,
and the references listed above is far from complete.

The molecular dynamics (MD) simulation has been proved to be useful in
understanding the behavior of a protein (or a short peptide segment)
in both static and oscillatory electric
fields~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB).
And then the temperature controlled simulation are performed for from
a few to tens of nanoseconds, during which the interested
conformational changes can be observed. Along the trajectories, the
secondary structure, root mean square displacement (RMSD), the dipole
moment and some shape parameters are calculated and investigated by
either plotted as a function of time, or doing time average over the
whole simulation. This approach release useful information on how the
secondary structure is changed by external electric field.  However,
by plotting the quantities against time, one may observe strong
fluctuation in plots, which hintered the quantitative and conclusive
analysis on what is happening during the conformational change, when
it happens and what is the strength. Although the statistical
uncertainty in the time-averaged observations are small, the
time-average itself is not well defined for a nonequilibrium process
such as relaxing process to the new conformations under static
electric field, and the dynamics under oscillating electric field.

The dynamical nonequilibrium molecular dynamics simulation (D-NEMD)
was firstly developed by G. Ciccotti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought}, and
recently successfully applied to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. D-NEMD provides a way of evaluating
nonequilibrium observables by doing an ensemble of nonequilibrium MD
simulations.  This provides a promising way of analyzing the
conformational changing under a generally time-dependent electric
field. In the present paper, we study the minimal model peptide
segment -- alanine dipeptide. We take the approach of starting from a
single peptide rather than a more realistic chain, firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding all possible interplay between
different peptide segments along a chain, then construct the chain
nonequilibrium properties from the well-studied single peptides.
Secondly, the current computational resource available is not enough
to study peptide chains by D-NEMD, since one needs to simulate an
ensemble of MD trajectories (typically a few thousands), so it
requires much higher computational cost than producing a single MD
trajectory.

In this paper, we firstly describe the idea of the D-NEMD, then
observables helping us the analyze the nonequilibrium properties
(i.e., the dipole moment, time-dependent properties of being a certain
conformation, the nonequilibrium transition properties and the net
property flux) are introduced. The numerical results of alanine
dipeptide under static and sine-shaped oscillating electric field are
presented. By calculating the nonequilibrium properties from the D-NEMD
simulations, we are able to answer from which to which the conformation
changes, when do these changes happens, and how strong are the changes.
As far as we know, our simulation is the first D-NEMD simulation
on the electric field effect on alanine dipeptide, and provides new
information helping people better understand the nonequilibrium
effect of electric field on the protein conformation.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
Here we fastly review the approach of performing nonequilibrium
molecular dynamics that was initiated by Giovanni Ciccoti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
We denote the macroscopic nonequilibrium observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
where $\hat O (\vect x)$ is the microscopic observable, 
which is measured at the phase space position $\vect x$,
comparing with the macroscopic observable that is not phase space independent.
In this work, we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known, and is equal to the equilibrium distribution
without any electric field.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes the inner produce in the
phase space.  We assume the dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x)$, where $\mathcal H$ is the Hamiltonian, and $J$ is
the symplectic matrix, then the Liouville equation of the probability
distribution is
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
The Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = e^{-iL(t)} \rho(\vect x, 0)$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by $\hat O(\vect x(t)) = e^{iL(t)} O(\vect x, 0)$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), e^{-iL(t)} \rho(\vect x, 0)\rangle
  = \langle e^{iL(t)}\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the initial
probability distribution is the equilibrium distribution, an
explicit translation of Eq.~\eqref{eqn:tmp4} is that the
nonequilibrium observable is equal to the ensemble average of
microscopic observable measured along trajectory $\vect x(t)$, the
initial configuration of which is subject to the \emph{equilibrium}
distribution. In practice, we firstly run an equilibrium MD simulation
to generate a sampling of configurations that are subject to the
equilibrium distribution. And then by using these configurations as
initial configuration, the Hamiltonian dynamics are integrated until
time $t$. These trajectories are also called \emph{branching
  trajectories} in this paper. Finally, the macroscopic observable is
estimated by averaging the microscopic observable measured at the end
points of the trajectories.

@@

The aforementioned nonequilibrium molecular dynamics simulation algorithm
does not describe how the boundary of the system should be set up.
The very naive idea of doing the simulation
is to simulate a infinitely large isolated system, or any large enough
isolated subsystem of it. In practice, it is both impossible and unnecessary to
do such a simulation, due to the unaffordable computational expense.
Therefore, proper boundary condition should be supplied
to truncated the simulation region. It is only possible when we
have a clear knowledge on
on how the nonequilibrium system is set up,
and what are the nonequilibrium observables to measure.

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig/thermostat.eps}
  \caption{Schematic plot of the truncated nonequilibrium subsystem.
    The white part is the subsystem and the gray part is the boundary
    condition, implemented by the thermostat and barostat mimicing
    the infinitely large universe.}
  \label{fig:tmp2}
\end{figure}

@@

\subsection{System settings and simulation protocol}
\begin{figure}
  \centering
  % \includegraphics[width=0.3\textwidth]{fig.ala/ele.field/field.eps}
  \includegraphics[]{fig.ala/field.dipole/fig-field-dipol-conf.eps}
  % \\
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/a1-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/a2-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/b1-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/b2-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/c-1.eps}
  \caption{The magnitude of the external electric field, and the
    x-component of the dipole moment of the alanine molecule as a
    function of time. The dashed lines are the strengths of the
    external electric field, while the solid lines are the dipole
    moments. Three test cases are indicated by different colors.}
  \label{fig:tmp3}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fig.ala/graphs/graph-t0010.eps}
  \caption{The schematic plot of the main probability flux among the metastable
    states on the Ramachandran histogram
    for periodically oscillating EF, $T=10$~ps. The thickness 
    of the arrows approximately presents the strength of the flux.
  }
  \label{fig:tmp11}
\end{figure}

>>>>

\title{Exploring Molecular Conformations and Related Time Scales via Nonequilibrium Molecular Dynamics Simulation}
% \title{The non-equilibrium molecular dynamics simulation of the alanine dipeptide under electric field}
\author{Han Wang}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin, Germany}
\author{Giovanni Ciccotti}
\affiliation{Department of Physics, University of Roma "La Sapienza", Italy}
\author{Luigi Delle Site}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

The possible effect of the electric field (EF) on the conformation of
proteins gains increasing attention recently, because it is closely
related to the health problem of exposing human tissue in the
electromagnetic radiation. Despite the importance of this problem, a
complete understanding of how the EF influence the  protein
is still lacking, therefore, a considerable number of
experimental~\cite{bohr2000microwave, bohr2000microwave-1,
  dePomerai2000cell, inskip2001cellular, mancinelli2004non} and
theoretical~\cite{budi2005electric, budi2007effect,
  budi2008comparative, toschi2008effects, astrakas2011electric,
  astrakas2012structural, damm2012can, starzyk2013proteins,
  english2009nonequilibrium, solomentsev2012effects} studies have been devoted to this topic,
and the references listed above is far from complete.

The molecular dynamics (MD) simulation has been proved to be useful in
understanding the behavior of a protein (or a short peptide segment)
in both static and oscillatory EF~\cite{budi2005electric, budi2007effect, budi2008comparative,
  toschi2008effects, astrakas2011electric, astrakas2012structural,
  damm2012can, starzyk2013proteins, english2009nonequilibrium,
  solomentsev2012effects}. In these studies, the initial
configurations are usually taken from the Protein Data Bank (PDB).
And then the temperature controlled simulation are performed for 
a few to tens of nanoseconds, during which the interested
conformational changes can be observed. Along the trajectories, the
secondary structure, root mean square displacement (RMSD), the dipole
moment and some shape parameters are calculated and investigated by
either plotted as a function of time, or doing time average over the
whole simulation. This approach release useful information on how the
secondary structure is changed by external  EF.  However,
by plotting the quantities against time, one may observe strong
fluctuation in plots, which hintered the quantitative and conclusive
analysis on what is the trend of the conformational change, when
it happens and what is the strength.
Although the time average reduces the statistical uncertainty,
it is indeed not well defined for a nonequilibrium process
such as relaxing process to the new conformations under a static
EF, or the dynamics under an oscillating EF.

The dynamical nonequilibrium molecular dynamics simulation (D-NEMD)
was firstly developed by G. Ciccotti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought}, and
recently successfully applied to the study of
hydrodynamics~\cite{orlandini2011hydrodynamics,
  orlandini2011hydrodynamics-01}. D-NEMD provides a way of evaluating
nonequilibrium observables by doing an ensemble of nonequilibrium MD
simulations.  It can be used for analyzing the
conformational changing under a generally time-dependent EF.
In the present paper, we study the minimal model peptide
segment -- alanine dipeptide. We take the approach of starting from a
single peptide piece rather than a more realistic chain, firstly because it
is easier to draw clear conclusions on the nonequilibrium properties
of a single peptide by excluding all possible interplay between
different peptide segments along a chain.
Secondly, the current computational resource available is not enough
to study peptide chains by D-NEMD, since one needs to simulate an
ensemble of MD trajectories (typically a few thousands), so it
requires much higher computational cost than producing a single
equilibrium MD trajectory.

In this paper, we firstly describe the idea of the D-NEMD, then
introduce the
observables for analyzing the nonequilibrium properties
(i.e., the dipole moment, time-dependent properties of being a certain
conformation and the net
probability flux). The numerical results of alanine
dipeptide under static and sine-shaped oscillating EF are
presented. By calculating the nonequilibrium properties from the D-NEMD
simulations, we are able to answer how the peptide
changes the conformation in a EF, when do these changes happens,
and how strong are they.
As far as we know, our simulation is the first D-NEMD simulation
of alanine dipeptide under EF, and provides new
information helping people better understand the nonequilibrium
effect of EF on the protein conformation.

\section{Methodology}
\subsection{Nonequilibrium molecular dynamics simulation}
Here we fastly review the approach of performing dynamical nonequilibrium
molecular dynamics (D-NEMD) that was initiated by Giovanni Ciccoti and his
colleagues~\cite{ciccotti1975direct, ciccotti1979thought,
  orlandini2011hydrodynamics, orlandini2011hydrodynamics-01}.
We denote the macroscopic nonequilibrium observable by $O(t)$. If at $t$
the configurational probability distribution is $\rho(\vect x, t)$, where
$\vect x$ is the phase space variable, then the observable can be
expressed by
\begin{align}\label{eqn:tmp1}
  O(t) = \int d\vect x\, \hat O(\vect x)\rho(\vect x, t)  = \langle \hat O(\vect x), \rho(\vect x, t)\rangle,
\end{align}
where $\hat O (\vect x)$ is the microscopic observable, 
which is measured at the phase space position $\vect x$,
comparing with the macroscopic observable that is not phase space independent.
In this work, we always assume that the initial probability distribution
$\rho(\vect x, 0)$ is known, and is equal to the equilibrium distribution
without any electric field.
The bracket on the right hand side of Eq.~\eqref{eqn:tmp1} denotes the inner produce in the
phase space.  We assume the dynamics of the system is governed by the
Hamiltonian equation, i.e. $\dot {\vect x} = J \cdot \nabla_{\vect x}
\mathcal H(\vect x)$, where $\mathcal H$ is the Hamiltonian, and $J$ is
the symplectic matrix, then the Liouville equation of the probability
distribution is
\begin{align}\label{eqn:tmp2}
  \frac{\partial \rho(\vect x, t)}{\partial t} = - iL(t) \rho(\vect x, t),
\end{align}
where $iL(t) = \{\cdot, \mathcal H\}$ is the Liouville operator.
The Eq.~\eqref{eqn:tmp2}
can be formally solved by $\rho(\vect x, t) = e^{-iL(t)} \rho(\vect x, 0)$.
On the other hand
\begin{align}
  \frac{d \hat O(\vect x(t))}{dt} = \nabla_{\vect x}\hat O\cdot \dot{\vect x}
  = \nabla_{\vect x}\hat O\cdot J\cdot \nabla_{\vect x}\mathcal H
  = iL(t) \hat O (\vect x(t))
\end{align}
This equation can be formally solved by $\hat O(\vect x(t)) = e^{iL(t)} O(\vect x, 0)$, therefore,
\begin{align}\nonumber
  O(t) & = \langle \hat O(\vect x), \rho(\vect x, t)\rangle
  = \langle \hat O(\vect x), e^{-iL(t)} \rho(\vect x, 0)\rangle
  = \langle e^{iL(t)}\hat O(\vect x), \rho(\vect x, 0)\rangle\\\label{eqn:tmp4}
  &= \langle \hat O(\vect x(t)), \rho(\vect x, 0)\rangle
\end{align}
Since we assume that the initial
probability distribution is the equilibrium distribution, an
explicit translation of Eq.~\eqref{eqn:tmp4} is that the
nonequilibrium observable is equal to the ensemble average of
microscopic observable measured along trajectory $\vect x(t)$, the
initial configuration of which is subject to the \emph{equilibrium}
distribution. In practice, we firstly run an equilibrium MD simulation
to generate a sampling of configurations that are subject to the
equilibrium distribution. And then by using these configurations as
initial configuration, the Hamiltonian dynamics are integrated until
time $t$. These trajectories are also called \emph{branching
  trajectories} in this paper. Finally, the macroscopic observable is
estimated by averaging the microscopic observable measured at the end
points of the trajectories.

@@

The aforementioned D-NEMD simulation algorithm
does not describe how the boundary of the system should be set up.
The very naive idea of doing the simulation
is to simulate a infinitely large isolated system, or any large enough
isolated subsystem of it. In practice, it is both impossible and unnecessary to
do such a simulation, due to the unaffordable computational expense.
Therefore, proper boundary condition should be supplied
to truncated the simulation region. It is only possible when we
have a clear knowledge on
on how the nonequilibrium system is set up,
and what are the nonequilibrium observables to measure.

\begin{figure}
  \centering
  \includegraphics[width=0.3\textwidth]{fig/thermostat.eps}
  \caption{Schematic plot of the truncated nonequilibrium subsystem.
    The white part is the Hamiltonian dynamics subsystem
    and the gray part is thermostating region. The whole system
    is subject to the periodic boundary condition.}
  \label{fig:tmp2}
\end{figure}

@@

\subsection{System settings and simulation protocol}
\begin{figure}
  \centering
  % \includegraphics[width=0.3\textwidth]{fig.ala/ele.field/field.eps}
  \includegraphics[]{fig.ala/field.dipole/fig-field-dipol-conf.eps}
  % \\
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/a1-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/a2-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/b1-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/b2-1.eps}
  % \includegraphics[width=0.18\textwidth]{fig.ala/confs/ext.mode1.010.Ex.01.00.t1000ps.recheck/c-1.eps}
  \caption{The magnitude of the external electric field, and the
    $x$-component of the dipole moment of the alanine molecule as a
    function of time. The dashed line presents the strengths of the
    external electric field, while the solid line presents the dipole
    moments. }
  \label{fig:tmp3}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fig.ala/graphs/graph-t0010-1.eps}
  \caption{The schematic plot of the main probability flux among the metastable
    states on the Ramachandran histogram
    for periodically oscillating EF, $T=10$~ps. The thickness 
    of the arrows approximately presents the strength of the flux.
  }
  \label{fig:tmp11}
\end{figure}
